---
title: "benchmarking_metrics"
output: html_document
date: "2024-10-31"
---

## Benchmarking metrics

We considered 3 different key metrics: the root-mean-squared error (RMSE), the Jensen--Shannon divergence (JSD) and the Spearman's rank correlation coefficient (ρ {\displaystyle \rho }). To create an overall benchmarking score against which to compare the deconvolution tools, we min-max scaled the metrics and computed the geometric mean of the three metrics to obtain the final benchmarking scores. Finally, we ranked the tools based on these scores.

## Import the libraries

```{r}
library(philentropy)
library(LaplacesDemon)
library(Metrics)
library(dplyr)
library(funkyheatmap)
library(tidyverse)
library(stringr)
library(funkyheatmap)
library(pROC)
```

## Import the dataset and filter it

```{r setup}
knitr::opts_knit$set(root.dir = "C:/Users/Sofie/OneDrive - UGent/Documents/Projects/DecoNFlow_benchmarking")
```


```{r}
# Combined all the results together
res_limma <- read.csv("data/Results_20M_limma.csv",row.names = 1)
res_dmrfinder <- read.csv("data/Results_20M_DMRfinder.csv",row.names = 1)
res_wgbstools <- read.csv("data/Results_20M_wgbstools.csv",row.names = 1)
res_cibersort_dmrfinder <- read.csv("data/Results_20M_DMRfinder_cibersort.csv",row.names = 1)
bench <- rbind(res_dmrfinder,res_limma,res_wgbstools,res_cibersort_dmrfinder)

# Combined results file with samples metadata for the expected tumoral fraction
metadata <- read.csv("data/SamplesMetadata.csv",sep = "\t")[,c("Sample","Exp.nbl")]
colnames(metadata) <- c("sample","expected_fraction")
bench <- as.data.frame(merge(bench,metadata,by="sample"))
# Use only the unbiased dataset
bench <- subset(bench,bench$reference=="reference_11healthy_9nbl" &
                bench$expected_fraction %in% c(0,0.0001,0.001,0.003,0.007,0.01,0.025,0.05,0.1,0.25,0.5))
bench$nbl <- round(bench$nbl,4)
```


```{r}
# Remove spaces at the beginning and end of the strings
bench$sample <- str_trim(bench$sample, side = c("both", "left", "right"))
bench$tool <- str_trim(bench$tool, side = c("both", "left", "right"))
# Remove the string due to the bam file format
bench$sample <- bench$sample %>% str_replace("_R1_001_val_1_bismark_bt2_pe", "")
bench <- as.data.frame(unique(bench))
```

## Plot the results

### Specify the metrics to use

For the 0% expected tumoral fraction, using RMSE can be problematic if you want to normalize it to make comparison per different tumor fractions possible. It's not possible to use MAPE either, since dividing by zero generates Inf values. That's the reason why we will use three different metrics: Matthew's correlation coefficient (MCC) for 0% fraction (we will treat this more as a "classification" problem), RMSE for fractions >0% and Spearman's rank correlation coefficient for all the fractions. We will therefore have 3 metrics that can be combined with a weighted linear combination, where the weights are represented by the proportion of samples used to compute each metric.

```{r}
# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Normalized RMSE (NRMSE)
nrmse <- function(actual, predicted) {
    round(sqrt(mean((actual - predicted))^2)/mean(actual),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman")
}
# MCC
mcc <- function(y_true,y_pred) {
  TP <- as.numeric(sum(y_true > 0 & y_pred > 0 ))
  TN <- as.numeric(sum(y_true == 0 & y_pred == 0))
  FP <- as.numeric(sum(y_true == 0 & y_pred > 0))
  FN <- as.numeric(sum(y_true > 0 & y_pred == 0))
  
  numerator <- (TP * TN) - (FP * FN)
  denominator <- sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))
  if (denominator == 0) return(0)
  return(numerator / denominator)
}
aupr.obj <- function(true_labels, predicted_scores) {
  true_labels[true_labels>0] <- 1

  # Calculate precision-recall curve and AUPR
  pr <- pr.curve(scores.class0 = predicted_scores, weights.class0 = true_labels, curve = T)
  
  return(pr)
}
roc.obj <- function(true_labels, predicted_scores) {
  true_labels[true_labels>0] <- 1
  roc_obj <- roc(true_labels, predicted_scores)
  return(roc_obj)
}
``` 

### Specify the output path

```{r}
path <- "C:/Users/Sofie/OneDrive - UGent/Documents/Projects/DecoNFlow_benchmarking/plots/"
```

### Plot boxplots of the predictions for each tumoral fraction > 0%

```{r}
# Create a directory to save the plots (optional)
output_dir <- paste0(path,"boxplots_tools_rmse")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of expected_fraction and remove the 0%
unique_fractions <- unique(bench$expected_fraction)
unique_fractions <- subset(unique_fractions,unique_fractions!=0)

# Loop over each expected_fraction and create/save the plot
for (fraction in unique_fractions) {
  
  # Filter for the specific fraction
  plot_data <- bench %>%
    filter(expected_fraction == fraction & tool!="Methyl_Resolver") %>%
    group_by(DMRtool)
  
  # Rank the tools by median difference
  median_diff <- bench %>% 
    filter(bench$expected_fraction==fraction) %>%
    group_by(tool,DMRtool) %>% 
    summarise(Diff=abs(median(expected_fraction)-median(nbl))) %>%
    group_by(tool) %>%
    summarise(Mean=mean(Diff, na.rm = T)) %>%
    arrange(Mean)
  
  # Step 2: Reorder the tools globally
  plot_data <- plot_data %>%
    mutate(tool = factor(tool, levels = median_diff$tool))
  
  p <- plot_data %>%
        group_by(tool) %>%
        ggplot(aes(x=tool,y=nbl,color=DMRtool)) +
        geom_boxplot() +
        geom_hline(yintercept = fraction,color = "red", linetype = "dashed") +
        labs(
          title = paste("Expected Fraction:", fraction),
          x = "",
          y = "Tumoral fraction"
        ) +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1)
          )

  # Save the plot as an image
  output_file <- file.path(output_dir, paste0("boxplots_tools_nomethylresolv_fraction_", fraction, ".png"))
  ggsave(output_file, plot = p, width = 8, height = 6)
}
1
```


### Plot each tool performance (NRMSE) per tumoral fractions >0%

We will use the normalized RMSE (NRMSE) in order to make the score size-free: otherwise, increasing the expected value will also determine an increase in RMSE, giving a (wrong) perception that tools perform worse when increasing the tumoral fraction.

```{r}

# Create a directory to save the plots (optional)
output_dir <- paste0(path,"tools_vs_fractions_nrmse")
if (!dir.exists(output_dir)) dir.create(output_dir)

for (dec_tool in unique(bench$tool)) {
  
    # Filter for the specific DMR selection tool
    plot_data <- bench %>%
      filter(bench$tool==dec_tool & bench$expected_fraction!=0) %>%
      group_by(DMRtool,expected_fraction) %>%  # Group by tool and tumoral fraction
      summarize(RMSE = nrmse(expected_fraction,nbl), .groups = "drop")  # Calculate mean RMSE
    
    # Plot the heatmap
    p <- ggplot(plot_data, aes(x = factor(expected_fraction), y = RMSE, color = DMRtool, shape = DMRtool)) +
      geom_point(size = 3, alpha = 0.8) +
      labs(
          title = paste("RMSE:", dec_tool),
          x = "Expected fraction",
          y = "RMSE"
      ) +
      theme(
          axis.text.x = element_text(angle = 45, hjust = 1)
      ) +
      scale_color_manual(values=c("limma"="#F8766D","wgbs_tools"="#00BA38","DMRfinder"="#619CFF"))
    
      # Save the plot as an image
      output_file <- file.path(output_dir, paste0(dec_tool,"_vs_fractions_rmse", ".png"))
      ggsave(output_file, plot = p, width = 8, height = 6)
}


```

### Plot tool ranks based on RMSE per tumoral fraction

```{r}
# Create a directory to save the plots
output_dir <- paste0(path,"ranked_tools_rmse")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of expected_fraction and exclude the 0
unique_fractions <- unique(bench$expected_fraction)
unique_fractions <- subset(unique_fractions,unique_fractions!=0)

# Loop over each expected_fraction and create/save the plot
for (fraction in unique_fractions) {
  # Filter for the specific fraction
  plot_data <- bench %>%
    filter(expected_fraction == fraction & tool!="Methyl_Resolver" ) %>%
    group_by(tool,DMRtool) %>% 
    summarise(RMSE = rmse(expected_fraction,nbl))
  
  # Rank the tools by mean RMSE across DMR tools
  median_diff <- bench %>% 
    filter(bench$expected_fraction==fraction) %>%
    group_by(tool,DMRtool) %>% 
    summarise(RMSE=rmse(expected_fraction,nbl)) %>%
    group_by(tool) %>%
    summarise(Mean=mean(RMSE, na.rm = T)) %>%
    arrange(Mean)
  
  # Reorder the tools globally
  plot_data <- plot_data %>%
    mutate(tool = factor(tool, levels = median_diff$tool))
  
  p <- plot_data %>%
        ggplot(aes(x=RMSE,y=fct_reorder(tool, RMSE),color=DMRtool, shape = DMRtool)) +
        geom_point(size = 3,alpha=0.8) +
        labs(
          title = paste("RMSE vs Tool (Expected Fraction:", fraction, ")"),
          x = "RMSE",
          y = ""
        )
  # Save the plot as an image
  output_file <- file.path(output_dir, paste0("ranking_tools_fraction_nomethylresolver_", fraction, ".png"))
  ggsave(output_file, plot = p, width = 8, height = 6)
}
```


### Plot AUC-ROC of tools at 4 low tumoral fractions

```{r}
# Create a directory to save the plots (optional)
output_dir <- paste0(path,"aucroc_tumoral_fractions")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Get 4 fractions: 0.0001, 0.001, 0.01, 0.05
aucroc_data <- data.frame()
fractions <- c(0.0001,0.001,0.01,0.05)
for (fraction in unique(fractions)) {
  df <- bench %>%
    filter(expected_fraction %in% c(0,fraction) & DMRtool == "DMRfinder") # & tool == 'CelFIE')
  for (deconv in unique(df$tool)) {
    filt_df <- df %>%
      filter(tool==deconv)
    roc_curve <- roc.obj(filt_df$expected_fraction,filt_df$nbl)
    tmp <- data.frame(
      fpr = 1-rev(roc_curve$specificities),  # False Positive Rate
      tpr = rev(roc_curve$sensitivities),  # True Positive Rate
      thresholds = rev(roc_curve$thresholds),
      auc = rev(roc_curve$auc),
      fraction = fraction,
      tool = deconv)  # Thresholds
    aucroc_data <- rbind(aucroc_data,tmp)
  }
}
```


```{r}
p <- ggplot(aucroc_data, aes(x = fpr, y = tpr, color = as.factor(fraction))) +
  geom_line(size = 1) +
  facet_wrap(~ tool, scales = "free", ncol = 5) +
  geom_point(aes(x = 0, y = auc, color = as.factor(fraction)),shape = 1,stroke=1.5,size=2,show.legend = F) +
  labs(
    title = "AUC-ROC at different tumoral fractions UXM",
    x = "FPR",
    y = "TPR",
    color = "Tumoral fraction"
  ) +
  theme(
    text = element_text(size = 12),
    strip.text = element_text(size = 10),
    legend.position = "bottom"
  )

# Save the plot as an image
output_file <- file.path(output_dir, paste0("aucroc_tumor_fractions_dmrfinder.png"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

```{r}
# Create the plot without facet_wrap
p <- ggplot(aucroc_data, aes(x = fpr, y = tpr, color = as.factor(fraction))) +
  geom_line(size = 1) +
  geom_point(aes(x = 0, y = auc, color = as.factor(fraction)), shape = 1, stroke = 1.5, size = 2, show.legend = FALSE) +
  labs(
    title = "AUC-ROC at different tumoral fractions",
    x = "FPR",
    y = "TPR",
    color = "Tumoral fraction",
    linetype = "Tool"
  ) +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )
p
```

### Plot final ranks

Use the MCC to evaluate how the tools perform when predicting presence/absence of tumor fraction. Then use the RMSE on the tumoral fractions >0% and the SCC on all tumoral fractions. Normalize all the metrics and integrate them using a linear, wighted combination. 

#### MCC to evaluate tools performance in presence/absence of tumor

```{r}
# Create a directory to save the plots (optional)
# output_dir <- paste0(path,"final_ranks")
#if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of expected_fraction and exclude the 0
#unique_fractions <- unique(bench$expected_fraction)
# Compute the metrics
# classif_performance <- bench %>%
#   group_by(tool,DMRtool) %>%
#   filter(!is.na(nbl)) %>%
#   summarize(MCC=mcc(expected_fraction,nbl))
```

#### AUC-ROC to evaluate tools performance in presence/absence of tumor

```{r}
# Create a directory to save the plots (optional)
output_dir <- paste0(path,"final_ranks")
if (!dir.exists(output_dir)) dir.create(output_dir)

aucroc_data <- data.frame()
fractions_auc <- unique(bench[bench$expected_fraction!=0,"expected_fraction"])
for (fraction in fractions_auc) {
  df <- bench %>%
    filter(expected_fraction %in% c(0,fraction))
  for (deconv in unique(df$tool)) {
    filt_df <- df %>%
      filter(tool==deconv)
    for (dmrtool in unique(filt_df$DMRtool)) {
      fin_df <- filt_df %>%
        filter(DMRtool==dmrtool)
      
      # Check input before calling `roc.obj`
      print(paste("Processing: Fraction =", fraction, 
                  ", Tool =", deconv, 
                  ", DMR Tool =", dmrtool))
      print("Unique values in expected_fraction (true labels):")
      print(unique(fin_df$expected_fraction))
      
      print("Checking nbl (predicted scores):")
      print(range(fin_df$nbl, na.rm = TRUE))
      
      roc_curve <- roc.obj(fin_df$expected_fraction,fin_df$nbl)
      tmp <- data.frame(
        fpr = 1-rev(roc_curve$specificities),  # False Positive Rate
        tpr = rev(roc_curve$sensitivities),  # True Positive Rate
        thresholds = rev(roc_curve$thresholds),
        auc = rev(roc_curve$auc),
        fraction = fraction,
        tool = deconv,
        DMRtool = dmrtool)
      aucroc_data <- rbind(aucroc_data,tmp)
    }
  }
}

classif_performance_auc <- aucroc_data %>%
  group_by(fraction,tool,DMRtool) %>%
  summarize(AUC=mean(auc)) %>%
  group_by(tool,DMRtool) %>%
  summarize(meanAUC=mean(AUC))

```

#### RMSE of tools on fractions > 0%

```{r}
# Create a directory to save the plots (optional)
# output_dir <- paste0(path,"final_ranks")
# if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of expected_fraction and exclude the 0
unique_fractions <- unique(bench$expected_fraction)

# Compute the metrics
nonzero_fraction <- bench %>%
  filter(expected_fraction!=0) %>%
  group_by(tool,DMRtool) %>%
  summarize(RMSE=rmse(expected_fraction,nbl))
```

#### Spearman's rank Correlation Coefficient on all the fractions

```{r}
# Create a directory to save the plots (optional)
# output_dir <- paste0(path,"final_ranks")
# if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of expected_fraction 
unique_fractions <- unique(bench$expected_fraction)

# Compute the metrics
all_fractions <- bench %>%
  group_by(tool,DMRtool) %>%
  summarize(SCC=scc(expected_fraction,nbl))
```

#### Plot general ranking of the tools based on RMSE, MCC, SCC, AUC and combined score

```{r}
# Create a directory to save the plots (optional)
output_dir <- paste0(path,"final_ranks")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of expected_fraction and exclude the 0
unique_fractions <- unique(bench$expected_fraction)

# Merge the metrics
merged_metrics <- merge(all_fractions,nonzero_fraction,by = c("tool","DMRtool"))
#merged_metrics <- merge(merged_metrics,classif_performance,by = c("tool","DMRtool"))
merged_metrics <- merge(merged_metrics,classif_performance_auc,by = c("tool","DMRtool"))

# Normalize all the metrics
normalized_list <- list()
for (selection in unique(merged_metrics$DMRtool)) {
  tmp <- na.omit(merged_metrics[merged_metrics$DMRtool==selection,])
  tmp$normSCC <- (tmp$SCC - min(tmp$SCC)) / (max(tmp$SCC)-min(tmp$SCC))
  tmp$normRMSE <- 1-(tmp$RMSE - min(tmp$RMSE)) / (max(tmp$RMSE) - min(tmp$RMSE))
  #tmp$normMCC <- (tmp$MCC - min(tmp$MCC)) / (max(tmp$MCC) - min(tmp$MCC))
  tmp$normAUC <- (tmp$meanAUC - min(tmp$meanAUC)) / (max(tmp$meanAUC)-min(tmp$meanAUC))
  # Append the normalized subset to the list
  normalized_list[[selection]] <- tmp[, c("tool", "DMRtool", "normSCC", "normRMSE","normAUC")]
}
# Combine all normalized subsets into a single data frame
normalized_df <- do.call(rbind, normalized_list)

# Create a combined metrics
nzeros <- nrow(bench[bench$expected_fraction==0,])
nnonzeros <- nrow(bench[bench$expected_fraction!=0,])
tot <- nzeros + nnonzeros
normalized_df$Score <- 
  normalized_df$normAUC +
  #normalized_df$normMCC +
  (nnonzeros/tot)*(normalized_df$normRMSE) + 
  normalized_df$normSCC

  
for (metric in c("normSCC","normRMSE","Score","normAUC")) {
    # Rank the tools by combined metric
    mean_score <- normalized_df %>% 
      group_by(tool) %>% 
      summarise(Mean=mean(.data[[metric]],na.rm = T)) %>%
      arrange(desc(Mean))
    # Reorder the tools globally
    normalized_df <- normalized_df %>%
      mutate(tool = factor(tool, levels = mean_score$tool))
    
    p <- ggplot(normalized_df, aes(y = tool, x = .data[[metric]], color = DMRtool, shape = DMRtool)) +
        geom_point(size = 3, alpha = 0.8) +
        labs(
            title = paste0("Tools ranked by mean ",metric),
            x = metric,
            y = ""
        ) +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1)
        ) +
        scale_color_manual(values=c("limma"="#F8766D","wgbs_tools"="#00BA38","DMRfinder"="#619CFF")) + theme_classic()
      
  # Save the plot as an image
  output_file <- file.path(output_dir, paste0("tools_vs_",metric,".png"))
  ggsave(output_file, plot = p, width = 8, height = 6) 
}

```


### Plot heatmap of tools performance (RMSE) per tumoral fraction

To plot the performance of each tool at different tumoral fractions, we will use the RMSE since the SCC cannot be used for fixed fractions (covariance is == 0)

```{r}
# Create a directory to save the plots (optional)
output_dir <- paste0(path,"heatmap_tools_rmse")
if (!dir.exists(output_dir)) dir.create(output_dir)

for (selection in unique(bench$DMRtool)) {
  
    # Filter for the specific DMR selection tool
    plot_data <- bench %>%
      filter(bench$DMRtool==selection & expected_fraction!=0 & bench$tool!="Methyl_Resolver") %>%
      group_by(tool,expected_fraction) %>%  # Group by tool and tumoral fraction
      summarize(RMSE = 1-mean(rmse(expected_fraction,nbl), na.rm = TRUE),
                .groups = "drop")
      
    # Rank the tools by median RMSE
    median_diff <- plot_data %>% 
      group_by(tool) %>% 
      summarise(Mean=abs(mean(RMSE))) %>%
      arrange(desc(Mean))
    
    # Reorder the tools globally
    plot_data <- plot_data %>%
      mutate(tool = factor(tool, levels = median_diff$tool))
    
    # Plot the heatmap
    p <- ggplot(plot_data, aes(x = tool, y = factor(expected_fraction), fill = RMSE)) +
      geom_tile() +  # Heatmap tiles
      geom_text(aes(label = round(RMSE, 4)), color = "white", size = 3) +  # Annotate RMSE
      scale_fill_viridis_c(end = 0.6) +  # Color gradient
      labs(
        title = paste0("Heatmap of Tumoral Fraction vs Tools (",selection,")"),
        x = "",
        y = "Tumoral Fraction"
      ) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        panel.grid = element_blank(),  # Remove gridlines for cleaner look
        
      )
    
      # Save the plot as an image
      output_file <- file.path(output_dir, paste0("heatmap_tools_wo_methylresolver", selection, ".png"))
      ggsave(output_file, plot = p, width = 8, height = 6)
}

library(plotly) 
```
