---
title: "benchmarking_metrics"
output: html_document
date: "2024-10-31"
---

## Benchmarking metrics

We considered 3 different key metrics: the root-mean-squared error (RMSE), the area under the curve (AUC-ROC) and the Spearman's rank correlation coefficient (ρ {\displaystyle \rho }). To create an overall benchmarking score against which to compare the deconvolution tools, we min-max scaled the metrics and computed the geometric mean of the three metrics to obtain the final benchmarking scores. Finally, we ranked the tools based on these scores.

## Import the libraries

```{r message=FALSE, warning=FALSE}
library(philentropy)
library(LaplacesDemon)
library(Metrics)
library(dplyr)
library(funkyheatmap)
library(tidyverse)
library(stringr)
library(funkyheatmap)
library(pROC)
```

## Import the datasets and filter it

Here, we will import all the results generated on the neuroblastoma vs healthy in-silico mixtures. A total number of 10 tumoral fractions will be considered. For each simulated tumoral fraction we have generated 60 replicates. Additionally, this scenario has been simulated with 4 different sequencing depth: an ideal scenario in the context of RRBS sequencing depth (20M of reads), and 3 other scenarios in which the initial samples have been downsampled: 15M, 10M and 5M or reads to estimate the deconvolution performance at lower sequencing depth.

```{r setup}
knitr::opts_knit$set(root.dir = "C:/Users/Sofie/OneDrive - UGent/Documents/Projects/DecoNFlow_Benchmarking/")
```

In this chunk we will merge all the results generated for each DMR selection tool at different sequencing depths. If you already have the final .csv file, you can skip this.

```{r}
# Define the directory containing the CSV files
dir_path <- "C:/Users/Sofie/OneDrive - UGent/Documents/Projects/DecoNFlow_Benchmarking/"

# Get a list of all CSV files in the directory
#colnames(bench)
csv_files <- list.files(paste0(dir_path,"results/"), pattern = "Results_.*", full.names = TRUE)

# Read and bind all CSV files
# combined_data <- do.call(rbind, lapply(csv_files, function(file) {
#   df <- read.csv(file, row.names = 1)
#   return(df)
# }))

# Save the combined file
#write.csv(combined_data, file = paste0(dir_path,"results/results_nbl_cfRRBS.csv"), row.names = FALSE)
```

The next step is to merge the benchmarking results file with the metadata file, where we will select 3 columns that are crucial for our analyses: sample, expected tumoral (Exp.nbl) fraction and the sequencing depth. Finally, as you can see from the combined_data file, there are 3 different reference datasets used for this analysis. For our experiment, we will select only one reference (11 healthy vs 9nbl) cause this reference is totally independent from the samples used to generate the in-silico simulated mixtures.

```{r}
metadata <- read.csv(paste0(dir_path,"files/samples_metadata_nbl_cfRRBS.csv"),sep = "\t")[,c("Sample","Exp.nbl","Depth")]
colnames(metadata) <- c("sample","expected_fraction","depth")

combined_data <- read.csv(paste0(dir_path,"results/results_nbl_cfRRBS.csv"))

bench <- as.data.frame(merge(combined_data,metadata,by="sample"))
bench <- subset(bench,bench$reference=="reference_11healthy_9nbl" &
                bench$expected_fraction %in% c(0,0.0001,0.001,0.003,0.007,0.01,0.025,0.05,0.1,0.25,0.5))
bench$nbl <- round(bench$nbl,4)
```

Here we perform some cleaning steps to avoid white space at the beginning and ending of the strings. We also remove the duplicates rows, which shouldn't be present anyway.

```{r}
bench$sample <- str_trim(bench$sample, side = c("both", "left", "right"))
bench$tool <- str_trim(bench$tool, side = c("both", "left", "right"))
#bench$sample <- bench$sample %>% str_replace("_R1_001_val_1_bismark_bt2_pe", "")
bench <- as.data.frame(unique(bench))

#write.csv(bench, file = paste0(dir_path,"bench.csv"), row.names = FALSE)

```

## Specify the metrics to use

The next step to do is to define the metrics that we will use to evaluate the tools. For the 0% expected tumoral fraction, using RMSE can be problematic if you want to normalize it to make comparison per different tumor fractions possible. It's not possible to use MAPE either, since dividing by zero generates Inf values. That's the reason why we will use three different metrics: AUC value for classification performance (we try to answer the following question: how good is a tool in predicting presence/absence of tumor?), RMSE for fractions \>0% (here we try to answer the following question: how good is a tool in estimating the tumor fractions in samples where we expect tumoral fraction?) and Spearman's rank correlation coefficient fractions \>= 0% (this will be used more as a general metric). We will therefore have 3 metrics that can be combined with a weighted linear combination, where the weights are represented by the proportion of samples used to compute each metric:

$$ 
\text{Score} = \text{AUC} \times 1 + \text{SCC} \times 1 + \text{RMSE} \times 0.87 
$$

Since RMSE is computed on a proportion of all the samples (specifically, only the samples with tumoral fraction \> 0% --\> 0.87), it will be weighted by this percentage. The normalized RMSE (NRMSE) will be used to compare tools performance across different tumoral fractions: if we don't normalize, increasing the tumoral fraction will result in an increase in RMSE just because the value are higher and not necessarly because the performance is worse. The AUC will be computed using the library pROC, so no need to write it.

```{r}
# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Normalized RMSE (NRMSE)
nrmse <- function(actual, predicted) {
    round(sqrt(mean((actual - predicted))^2)/mean(actual),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman")
}
roc.obj <- function(true_labels, predicted_scores) {
  true_labels[true_labels>0] <- 1
  roc_obj <- roc(true_labels, predicted_scores)
  return(roc_obj)
}

```

## Plot the benchmarking results

Now we are ready to plot the results for downstream interpretation.

### Plot boxplots of the predictions for each tumoral fraction \> 0%

The first visualization we are plotting are boxplots or the predicted tumoral fractions for each deconvolution tool across 3 different DMR selection tools. We will create one folder for each sequencing depth, and within each folder, one folder for mean and one folder for median.

```{r}
# Create the main directory to save the plots
output_dir <- paste0(dir_path, "plots/boxplots_tools_rmse")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of sequencing depth and expected fraction
unique_depths <- unique(bench$depth)
unique_approaches <- unique(bench$collapse_approach)
unique_fractions <- unique(bench$expected_fraction)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over each sequencing depth
for (seqdepth in unique_depths) {
  
  # Create a subdirectory for this depth (optional)
  depth_dir <- file.path(output_dir, paste0("depth_", seqdepth))
  if (!dir.exists(depth_dir)) dir.create(depth_dir)

  # Loop over collapse approach
  for (approach in unique_approaches) {
    
    # Create a subdirectory for this collapse approach (optional)
    approach_dir <- file.path(depth_dir, approach)
    if (!dir.exists(approach_dir)) dir.create(approach_dir)
    
      # Loop over each expected fraction
    for (fraction in unique_fractions) {
      
      # Filter for the specific depth and fraction
      plot_data <- bench %>%
        filter(depth == seqdepth & expected_fraction == fraction &
               collapse_approach == approach)
      
      # Rank the tools by median difference
      median_diff <- plot_data %>% 
        group_by(tool, DMRtool) %>% 
        summarise(Diff = abs(median(expected_fraction) - median(nbl)), .groups = "drop") %>%
        group_by(tool) %>%
        summarise(Mean = mean(Diff, na.rm = TRUE), .groups = "drop") %>%
        arrange(Mean)
      
      # Step 2: Reorder the tools globally
      plot_data <- plot_data %>%
        mutate(tool = factor(tool, levels = median_diff$tool))
      
      # Generate the boxplot
      p <- plot_data %>%
        group_by(tool) %>%
        ggplot(aes(x = tool, y = nbl, color = DMRtool)) +
        geom_boxplot() +
        geom_hline(yintercept = fraction, color = "red", linetype = "dashed") +
        labs(
          title = paste("Depth:", seqdepth, "| Approach:", approach, "| Expected Fraction:", fraction),
          x = "",
          y = "Tumoral fraction"
        ) +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1)
        )
      
      # Save the plot
      output_file <- file.path(approach_dir, paste0("boxplot_depth_", seqdepth, 
                                                    "_approach_", approach, 
                                                    "_fraction_", fraction, ".png"))
      ggsave(output_file, plot = p, width = 8, height = 6)
    }
  }
}
```

### Plot each tool performance (NRMSE) per tumoral fractions \>0%

Now it might be interesting how the deconvolutions tools perform when increasing (or decreasing) the tumoral fraction. That is, what is the deconvolution performance at each tumoral fraction. However, since RMSE is scale dependent, we need to normalize it. We will use the normalized RMSE (NRMSE) in order to make the score scale indipendent: otherwise, increasing the expected value will also determine an increase in RMSE, giving a (wrong) perception that tools perform worse when increasing the tumoral fraction. The NRMSE is calculated as following:

$$
\text{NRMSE} = \frac{\text{RMSE}}{\bar{y}}
$$

```{r}
# Create the main directory to save the plots
output_dir <- paste0(dir_path, "plots/tools_vs_fractions_nrmse")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of sequencing depth, collapse approach, and tools
unique_depths <- unique(bench$depth)
unique_approaches <- unique(bench$collapse_approach)
unique_tools <- unique(bench$tool)

# Loop over sequencing depth
for (seqdepth in unique_depths) {
  
  # Create a subdirectory for this depth (optional)
  depth_dir <- file.path(output_dir, paste0("depth_", seqdepth))
  if (!dir.exists(depth_dir)) dir.create(depth_dir)
  
  # Loop over collapse approach
  for (approach in unique_approaches) {
    
    # Create a subdirectory for this collapse approach (optional)
    approach_dir <- file.path(depth_dir, approach)
    if (!dir.exists(approach_dir)) dir.create(approach_dir)
    
    # Loop over each tool (dec_tool)
    for (dec_tool in unique_tools) {
      
      # Filter for the specific depth, collapse approach, and tool
      plot_data <- bench %>%
        filter(depth == seqdepth & collapse_approach == approach & 
               tool == dec_tool & expected_fraction != 0) %>%
        group_by(DMRtool, expected_fraction) %>%
        summarize(NRMSE = rmse(expected_fraction, nbl) / mean(expected_fraction), .groups = "drop")  # Calculate mean RMSE
      
      # Plot the heatmap
      p <- ggplot(plot_data, aes(x = factor(expected_fraction), y = NRMSE, color = DMRtool, shape = DMRtool)) +
        geom_point(size = 3, alpha = 0.8) +
        labs(
          title = paste("NRMSE:", dec_tool, "| Depth:", seqdepth, "| Approach:", approach),
          x = "Expected fraction",
          y = "NRMSE"
        ) +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1)
        ) +
        scale_color_manual(values = c("limma" = "#F8766D", "wgbs_tools" = "#00BA38", "DMRfinder" = "#619CFF"))
      
      # Save the plot as an image
      output_file <- file.path(approach_dir, paste0("NRMSE_", dec_tool, "_depth_", seqdepth, "_", approach, ".png"))
      ggsave(output_file, plot = p, width = 8, height = 6)
    }
  }
}


```


### Plot tool ranks based on RMSE per tumoral fraction

Here, we want to show the deconvolution tools perfomance at each tumoral fraction. Very similar to what we plot above, but in this case we are ranking for each tumoral fraction the tools based on the RMSE. Since in this analysis we are only looking at each tumoral fraction separately, there is no need to normalize the RMSE.


```{r}
# Create the main directory to save the plots
output_dir <- paste0(dir_path, "plots/ranked_tools_rmse")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Get unique values of sequencing depth, collapse approach, and expected_fraction
unique_depths <- unique(bench$depth)
unique_approaches <- unique(bench$collapse_approach)
unique_fractions <- unique(bench$expected_fraction)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over sequencing depth
for (seqdepth in unique_depths) {
  
  # Create a subdirectory for this depth (optional)
  depth_dir <- file.path(output_dir, paste0("depth_", seqdepth))
  if (!dir.exists(depth_dir)) dir.create(depth_dir)
  
  # Loop over collapse approach
  for (approach in unique_approaches) {
    
    # Create a subdirectory for this collapse approach (optional)
    approach_dir <- file.path(depth_dir, approach)
    if (!dir.exists(approach_dir)) dir.create(approach_dir)
    
    # Loop over each expected fraction
    for (fraction in unique_fractions) {
      
      # Filter for the specific depth, approach, and fraction
      plot_data <- bench %>%
        filter(depth == seqdepth & collapse_approach == approach & expected_fraction == fraction) %>%
        group_by(tool, DMRtool) %>% 
        summarise(RMSE = rmse(expected_fraction, nbl), .groups = "drop")
      
      # Rank the tools by mean RMSE across DMR tools
      median_diff <- plot_data %>%
        group_by(tool) %>%
        summarise(Mean = mean(RMSE, na.rm = TRUE)) %>%
        arrange(Mean)
      
      # Reorder the tools globally
      plot_data <- plot_data %>%
        mutate(tool = factor(tool, levels = median_diff$tool))
      
      # Generate the plot
      p <- ggplot(plot_data, aes(x = RMSE, y = tool, color = DMRtool, shape = DMRtool)) +
        geom_point(size = 3, alpha = 0.8) +
        labs(
          title = paste("RMSE vs Tool | Fraction:", fraction, "| Depth:", depth, "| Approach:", approach),
          x = "RMSE",
          y = ""
        )
      
      # Save the plot as an image
      output_file <- file.path(approach_dir, paste0("ranking_tools_fraction_", fraction, "_depth_", seqdepth, "_", approach, ".png"))
      ggsave(output_file, plot = p, width = 8, height = 6)
    }
  }
}
```

### Plot AUC-ROC of tools at 4 low tumoral fractions

Here we want to answer the question: how good are the deconvolution tools in predicting the presence/absence of the tumor at different tumoral fractions. Here we will investigate the AUC-ROC at 4, very low, tumoral fractions, that will be compared in pair-wise comparisons with 0% tumor mixtures. For each pair-wise comparison, the AUC-ROC will be computed.

```{r}
# Create the main directory to save the plots
output_dir <- paste0(dir_path, "plots/aucroc_tumoral_fractions")
if (!dir.exists(output_dir)) dir.create(output_dir)

# Define the target fractions
fractions <- c(0.0001, 0.001, 0.01, 0.05)

# Get unique sequencing depths, collapse approaches, and DMR tools
unique_depths <- unique(bench$depth)
unique_approaches <- unique(bench$collapse_approach)
unique_dmrtools <- unique(bench$DMRtool)
miss <- list()
# Loop over sequencing depth
for (seqdepth in unique_depths) {
  
  # Create a subdirectory for this depth (optional)
  depth_dir <- file.path(output_dir, paste0("depth_", seqdepth))
  if (!dir.exists(depth_dir)) dir.create(depth_dir)
  
  # Loop over collapse approach
  for (approach in unique_approaches) {
    
    # Create a subdirectory for this collapse approach (optional)
    approach_dir <- file.path(depth_dir, approach)
    if (!dir.exists(approach_dir)) dir.create(approach_dir)
    
    for (dmrtool in unique_dmrtools) {
      aucroc_data <- data.frame()

      for (deconv in unique(df$tool)) {

        for (fraction in fractions) {
          filt_df <- bench %>%
            filter(expected_fraction %in% c(0,fraction) & 
                   DMRtool == dmrtool &
                   collapse_approach == approach &
                   depth == seqdepth &
                   tool == deconv)
          # Skip if there are not exactly two levels in expected_fraction (i.e., both 0 and fraction)
          if (length(unique(filt_df$expected_fraction)) != 2) {
            miss <- c(miss,deconv)
            next
          }
          roc_curve <- roc.obj(filt_df$expected_fraction,filt_df$nbl)
          tmp <- data.frame(
            fpr = 1-rev(roc_curve$specificities),  # False Positive Rate
            tpr = rev(roc_curve$sensitivities),  # True Positive Rate
            thresholds = rev(roc_curve$thresholds),
            auc = rev(roc_curve$auc),
            fraction = fraction,
            tool = deconv)
          aucroc_data <- rbind(aucroc_data,tmp)
        }
      }
      p <- ggplot(aucroc_data, aes(x = fpr, y = tpr, color = as.factor(fraction))) +
        geom_line(size = 1) +
        facet_wrap(~ tool, scales = "free", ncol = 5) +
        geom_point(aes(x = 0, y = auc, color = as.factor(fraction)),shape = 1,stroke=1.5,size=2,show.legend = F) +
        labs(
          title = "AUC-ROC at different tumoral fractions",
          x = "FPR",
          y = "TPR",
          color = "Tumoral fraction"
        ) +
        theme(
          text = element_text(size = 12),
          strip.text = element_text(size = 10),
          legend.position = "bottom"
        )
      
      # Save the plot as an image
      output_file <- file.path(approach_dir, paste0("auc", "_depth_", seqdepth, "_", approach, "_", dmrtool, ".png"))
      ggsave(output_file, plot = p, width = 8, height = 6)
    }
  }
}

```


### Plot final ranks

For now, we have only plotted the tools based on some metrics without making a ranking. In this section instead we will plot the tools ranked by the metrics we introduced at the beginning of this tutorial: AUC (presence/absence of tumor), RMSE (for >0%) and SCC (for all). In order to avoid unbalanced datasets when computing the AUC, each tumoral fraction will be compared to the 0% samples and AUC calculated. Finally, all the pairwise AUC will be averaged to extract the final AUC per tool.

#### AUC-ROC to evaluate tools performance in presence/absence of tumor

```{r message=FALSE, warning=FALSE}
# Get unique sequencing depths, collapse approaches, and DMR tools
unique_depths <- unique(bench$depth)
unique_approaches <- unique(bench$collapse_approach)
unique_dmrtools <- unique(bench$DMRtool)
aucroc_data <- data.frame()

# Loop over sequencing depth
for (seqdepth in unique_depths) {
  
  # Create a subdirectory for this depth (optional)
  depth_dir <- file.path(output_dir, paste0("depth_", seqdepth))
  if (!dir.exists(depth_dir)) dir.create(depth_dir)
  
  # Loop over collapse approach
  for (approach in unique_approaches) {
    
    # Create a subdirectory for this collapse approach (optional)
    approach_dir <- file.path(depth_dir, approach)
    if (!dir.exists(approach_dir)) dir.create(approach_dir)
    
    for (dmrtool in unique_dmrtools) {

      for (deconv in unique(df$tool)) {

        for (fraction in fractions) {
          filt_df <- bench %>%
            filter(expected_fraction %in% c(0,fraction) & 
                   DMRtool == dmrtool &
                   collapse_approach == approach &
                   depth == seqdepth &
                   tool == deconv)
          
          # Skip if df is empty
          if (length(unique(filt_df$expected_fraction)) != 2) {
            miss <- c(miss,deconv)
            next
          }
          
          roc_curve <- roc.obj(filt_df$expected_fraction,filt_df$nbl)
          tmp <- data.frame(
            fpr = 1-rev(roc_curve$specificities),  # False Positive Rate
            tpr = rev(roc_curve$sensitivities),  # True Positive Rate
            thresholds = rev(roc_curve$thresholds),
            auc = rev(roc_curve$auc),
            fraction = fraction,
            tool = deconv,
            DMRtool = dmrtool,
            collapse_approach = approach,
            depth = seqdepth)
          aucroc_data <- rbind(aucroc_data,tmp)
        }
      }
    }
  }
}
classif_performance_auc <- aucroc_data %>%
  group_by(fraction,tool,DMRtool,collapse_approach,depth) %>%
  summarize(AUC=mean(auc)) %>%
  group_by(tool,DMRtool,collapse_approach,depth) %>%
  summarize(meanAUC=mean(AUC))

```

#### RMSE of tools on fractions \> 0%

```{r}
# Get unique values of expected_fraction and exclude the 0
unique_fractions <- unique(bench$expected_fraction)

# Compute the metrics
nonzero_fraction <- bench %>%
  filter(expected_fraction!=0) %>%
  group_by(tool,DMRtool,collapse_approach,depth) %>%
  summarize(RMSE=rmse(expected_fraction,nbl))
```

#### Spearman's rank Correlation Coefficient on all the fractions

```{r}
# Get unique values of expected_fraction 
unique_fractions <- unique(bench$expected_fraction)

# Compute the metrics
all_fractions <- bench %>%
  group_by(tool,DMRtool,depth,collapse_approach) %>%
  summarize(SCC=scc(expected_fraction,nbl))
```

#### Plot general ranking of the tools based on RMSE, SCC, AUC and combined score

```{r}
# Get unique values of expected_fraction and exclude the 0
unique_fractions <- unique(bench$expected_fraction)
unique_depths <- unique(bench$depth)
unique_approaches <- unique(bench$collapse_approach)
unique_dmrtools <- unique(bench$DMRtool)

# Merge the metrics
merged_metrics <- merge(all_fractions,nonzero_fraction,by = c("tool","DMRtool","collapse_approach","depth"))
merged_metrics <- merge(merged_metrics,classif_performance_auc,by = c("tool","DMRtool","collapse_approach","depth"))

# Normalize all the metrics
normalized_list <- list()
miss <- list()
for (dmrtool in unique_dmrtools) {
  for (seqdepth in unique_depths) {
    for (approach in unique_approaches) {
      
      tmp <- merged_metrics %>%
              filter(DMRtool == dmrtool &
                   collapse_approach == approach &
                   depth == seqdepth)
      # Skip if df is empty
      if (dim(tmp)[1] == 0) {
        miss <- c(miss,deconv)
        next
      }
      tmp <- tmp %>%
        mutate(
          SCC = ifelse(is.na(SCC), 0, SCC),  # Replace NAs in SCC with 0
          RMSE = ifelse(is.na(RMSE), 1, RMSE),   # Replace NAs in RMSE with 1
          AUC = ifelse(is.na(meanAUC),0,meanAUC)  
        )
      tmp$normSCC <- (tmp$SCC - min(tmp$SCC)) / (max(tmp$SCC)-min(tmp$SCC))
      tmp$normRMSE <- 1-(tmp$RMSE - min(tmp$RMSE)) / (max(tmp$RMSE) - min(tmp$RMSE))
      tmp$normAUC <- (tmp$meanAUC - min(tmp$meanAUC)) / (max(tmp$meanAUC)-min(tmp$meanAUC))
      # Append the normalized subset to the list
      key <- paste(dmrtool, seqdepth, approach, sep = "_")
      normalized_list[[key]] <- tmp[, c("tool", "DMRtool", "collapse_approach","depth", "normSCC", "normRMSE","normAUC")]  
    }
  }
}
# Combine all normalized subsets into a single data frame
normalized_df <- do.call(rbind, normalized_list)

# Create a combined metrics
nzeros <- nrow(bench[bench$expected_fraction==0,])
nnonzeros <- nrow(bench[bench$expected_fraction!=0,])
tot <- nzeros + nnonzeros
normalized_df$Score <- 
  normalized_df$normAUC +
  (nnonzeros/tot)*(normalized_df$normRMSE) + 
  normalized_df$normSCC
```

Now we can plot the final ranks.

```{r}
# Create a directory to save the plots (optional)
output_dir <- paste0(dir_path,"plots/final_ranks")
if (!dir.exists(output_dir)) dir.create(output_dir)

for (metric in c("normSCC","normRMSE","Score","normAUC")) {
  
  for (seqdepth in unique_depths) {
    
    # Create a subdirectory for this depth (optional)
    depth_dir <- file.path(output_dir, paste0("depth_", seqdepth))
    if (!dir.exists(depth_dir)) dir.create(depth_dir)
    
    for (approach in unique_approaches) {
      
      # Create a subdirectory for this collapse approach (optional)
      approach_dir <- file.path(depth_dir, approach)
      if (!dir.exists(approach_dir)) dir.create(approach_dir)
      
      df <- normalized_df %>%
            filter(collapse_approach == approach &
                   depth == seqdepth)
      
      # Rank the tools by combined metric
      mean_score <- df %>% 
            group_by(tool) %>% 
            summarise(Mean=sum(.data[[metric]])/length(unique_dmrtools)) %>%
            arrange(Mean)

      # Reorder the tools globally
      df <- df %>%
        mutate(tool = factor(tool, levels = mean_score$tool))
      
      p <- ggplot(df, aes(y = tool, x = .data[[metric]], color = DMRtool, shape = DMRtool)) +
          geom_point(size = 3, alpha = 0.8) +
          labs(
              title = paste("Depth:", depth, "| Approach:", approach),
              x = metric,
              y = ""
          ) +
          theme(
              axis.text.x = element_text(angle = 45, hjust = 1)
          ) +
          scale_color_manual(values=c("limma"="#F8766D","wgbs_tools"="#00BA38","DMRfinder"="#619CFF"))
      
      # Save the plot as an image
      output_file <- file.path(approach_dir, paste0("rank_",metric,"_depth_", seqdepth, "_", approach, ".png"))
      ggsave(output_file, plot = p, width = 8, height = 6)
    }
  }
}
```


### Plot heatmap of tools performance (RMSE) per tumoral fraction

To plot the performance of each tool at different tumoral fractions, we will use the RMSE. We will plot an heatmap to have a more condensed overview.

```{r}
# Create a directory to save the plots (optional)
output_dir <- paste0(dir_path,"plots/heatmap_tools_rmse")
if (!dir.exists(output_dir)) dir.create(output_dir)
unique_fractions <- unique(bench$expected_fraction)
unique_depths <- unique(bench$depth)
unique_approaches <- unique(bench$collapse_approach)
unique_dmrtools <- unique(bench$DMRtool)

for (dmrtool in unique_dmrtools) {
  for (seqdepth in unique_depths) {
    
    # Create a subdirectory for this depth (optional)
    depth_dir <- file.path(output_dir, paste0("depth_", seqdepth))
    if (!dir.exists(depth_dir)) dir.create(depth_dir)
    
    for (approach in unique_approaches) {

      # Create a subdirectory for this collapse approach (optional)
      approach_dir <- file.path(depth_dir, approach)
      if (!dir.exists(approach_dir)) dir.create(approach_dir)
      
      # Filter for the specific DMR selection tool
      plot_data <- bench %>%
        filter(DMRtool==dmrtool & 
               expected_fraction!=0 &
               depth == seqdepth &
               collapse_approach == approach) %>%
        group_by(tool,expected_fraction) %>%  # Group by tool and tumoral fraction
        summarize(RMSE = rmse(expected_fraction, nbl), .groups = "drop")
        
      # Rank the tools by median RMSE
      median_diff <- plot_data %>% 
        group_by(tool) %>% 
        filter(expected_fraction==0.0001) %>%
        arrange(RMSE)
      
      # Reorder the tools globally
      plot_data <- plot_data %>%
        mutate(tool = factor(tool, levels = median_diff$tool))
      
      # Plot the heatmap
      p <- ggplot(plot_data, aes(x = tool, y = factor(expected_fraction), fill = RMSE)) +
        geom_tile() +  # Heatmap tiles
        geom_text(aes(label = round(RMSE, 5)), color = "white", size = 3) +  # Annotate RMSE
        scale_fill_viridis_c(end = 0.6) +  # Color gradient
        labs(
          title = paste0("Heatmap of Tumoral Fraction vs Tools (",dmrtool,")"),
          x = "",
          y = "Tumoral Fraction"
        ) +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
          panel.grid = element_blank()  # Remove gridlines for cleaner look
        )
      
        # Save the plot as an image
        output_file <- file.path(approach_dir, paste0("heatmap_tools_",dmrtool,"_depth_", seqdepth, "_", approach, ".png"))
        ggsave(output_file, plot = p, width = 8, height = 6)   
    }
  }
}

 
```

### Plot funkyheatmap

Prepare the data

```{r}
fraction_funkyheat <- bench %>%
  filter(expected_fraction!=0) %>%
  group_by(tool,expected_fraction,DMRtool) %>%
  summarize(RMSE=rmse(expected_fraction,nbl)) %>%
  mutate(
    fraction_class = cut(
      expected_fraction,
      breaks = c(-Inf, 0.001, 0.01, 0.1, 0.5, 1), # Define your breakpoints
      labels = c("<=0.001","<=0.01", "<=0.1", "<=0.5","1"), # Define labels for the classes
      right = TRUE # Interval includes the lower bound but not the upper
    )
  ) %>%
  group_by(fraction_class,tool) %>%
  summarize(RMSE_class=1-mean(RMSE),.groups = "drop") %>%
  pivot_wider(
    names_from = c(fraction_class),
    names_sep = "_",
    values_from = RMSE_class
  ) %>%
  mutate(across(where(is.numeric), 
                ~ (.-min(., na.rm = TRUE)) / (max(., na.rm = TRUE) - min(., na.rm = TRUE))))

dmrtool_funkyheat <- bench %>%
  group_by(DMRtool,tool) %>%
  summarize(RMSE=1-rmse(expected_fraction,nbl),.groups = "drop") %>%
  pivot_wider(
    names_from = c(DMRtool),
    names_sep = "_",
    values_from = RMSE
  ) %>%
  mutate(across(where(is.numeric), 
                ~ (.-min(., na.rm = TRUE)) / (max(., na.rm = TRUE) - min(., na.rm = TRUE))))

classif_funkyheat <- normalized_df %>%
  group_by(tool) %>%
  summarize(Classification=mean(normMCC))

metrics_funkyheat <- normalized_df %>% 
  group_by(tool) %>%
  summarize(Score=mean(Score),
            RMSE=mean(normRMSE),
            SCC=mean(normSCC),
            MCC=mean(normMCC)
            ) %>%
  mutate(Score=(Score-min(Score))/(max(Score)-min(Score)))

data_funkyheat <- merge(metrics_funkyheat[,c("tool","RMSE","SCC","MCC")],dmrtool_funkyheat,by=c("tool"))
data_funkyheat <- merge(data_funkyheat,fraction_funkyheat,by = c("tool"))
data_funkyheat <- merge(data_funkyheat,classif_funkyheat,by=c("tool"))
data_funkyheat <- merge(data_funkyheat,metrics_funkyheat[,c("tool","Score")],by=c("tool"))
data_funkyheat$Rank <- rank(-data_funkyheat$Score)
data_funkyheat <- data_funkyheat[order(data_funkyheat$Rank,decreasing = F),]
```

Prepare the column info necessary for the heatmap

```{r}
column_info <- tribble(
  ~id,            ~group,               ~name,         ~geom,        ~palette,    ~options,
  "tool",         "",                   "",            "text",       NA,          list(hjust = 0, width = 6),
  "RMSE",         "Per metric",         "1-RMSE",      "funkyrect",  "palette1",  lst(),
  "SCC",          "Per metric",         "SCC",         "funkyrect",  "palette1",  lst(),
  "MCC",          "Per metric",         "MCC",         "funkyrect",  "palette1",  lst(),
  "DMRfinder",    "Per DMR tool",       "DMRfinder",   "funkyrect",  "palette1",  lst(),
  "limma",        "Per DMR tool",       "limma",       "funkyrect",  "palette1",  lst(),
  "wgbs_tools",   "Per DMR tool",       "wgbs_tools",  "funkyrect",  "palette1",  lst(),
  "<=0.001",      "Per fraction",       "(0-0.001]",   "funkyrect",  "palette1",  lst(),
  "<=0.01",       "Per fraction",       "(0.001-0.01]","funkyrect",  "palette1",  lst(),
  "<=0.1",        "Per fraction",       "(0.01-0.1]",  "funkyrect",  "palette1",  lst(),
  "<=0.5",        "Per fraction",       "(0.1-0.5]",   "funkyrect",  "palette1",  lst(),
  "1",            "Per fraction",       "(0.5-1]",     "funkyrect",  "palette1",  lst(),
  "Score",        "Overall",            "Overall score","funkyrect", "palette2",  lst(),
  "Rank",         "Overall",            "Rank",        "rect",       "rank_rect", lst(legend = FALSE),
  "Rank",         "Overall",            "",            "text",       "rank_text", list(overlay=T),
)

palettes <- tribble(
  ~palette,    ~colours, 
  "rank_rect", rep("#DDAD4B",length(data_funkyheat$tool)),
  "rank_text", rep("black",length(data_funkyheat$tool)),
)
```

Plot and save it in a pdf

```{r}

g <- funky_heatmap(data_funkyheat, 
                   column_info = column_info,
                   palettes = palettes
                   ) 

# Create a directory to save the plots (optional)
output_dir <- "~/Desktop/plots/funkyheatmaps"
if (!dir.exists(output_dir)) dir.create(output_dir)
output_file <- file.path(output_dir, paste0("funkyheatmap.pdf"))
ggsave(output_file, g, width = g$width, height = g$height)
```
